Modelling Structured Societies: a Multi-relational Approach to Context Permeability

### 2015-06-21 doc/main.tex
overlapping network properties reviewed

        Since we construct scenarios where we overlap highly clustered k-regular networks and
        scale-free networks with node hubs, we are interested in the kind of properties that emerge
        from merging these structures -- and if/how they are comparable to the existing models
        --. The networks were produced by the \textit{b-have network library}
        \cite{Nunes:Software:11069}, exported to files, and analysed using the \textit{igraph R
          package} \cite{igraph2006}.

@@

% Overview
\noindent In this section, we analyse and discuss the results for different experiments with both the
context permeability and context switching models. We start by analysing the structural properties
that arise from combining different networks while maintaining the same set of network nodes. We then perform multiple experiments where we observe the impact of different model configurations in the convergence to consensus. We correlate the outcomes of our context permeability model simulations with the structural properties of the multi-network structures. We also explore how the new dynamics introduced by the context switching model, affect the consensus building process.

\subsection{Overlapping Network Properties}
\label{sec:network_properties}
\noindent In this first analysis, we investigate the properties of different network topologies that we use in our models. 
One of the parameters in some experiments is the number of networks in which the agents interact. Adding more networks means adding more connections. 
We show that what is important is not just the the number of connections, but the properties the resulting structure displays when these networks are merged. The way we look at the network properties is as follows.

Each network is generated in such a way that the node positions are randomised. This means that we can have multiple networks with the same topology and yet, each node has a different neighbourhood on each network. Also, neighbourhoods do not necessarily overlap due to the node shuffling. Adding more networks to the social space is not the same as creating a single network with twice the number of connections.

\noindent Consider figure \ref{fig:network_properties_merge_2_10regular}. We created two random k-regular networks with k=10 and merged them ignoring edges from common neighbours --in the simulation models this edge agglutination is not performed, we use it here for the purposes of network analysis. The common edges are highlighted in a different colour.

@@

The networks are drawn using the \textit{Kamada-Kawai} Layout \cite{Kamada19897} which regards the desirable ``geometric'' (Euclidean) distance between two vertices in the drawing as the ``graph theoretic'' distance between them in the corresponding graph. We can see by the results of the layout in figure \ref{fig:network_properties_merge_2_10regular}, that the distance between nodes in the network has decreased, this is also confirmed by the reduced average path length of the combined networks (see table \ref{tab:network_properties_merge_2_10regular}). To illustrate our point we analysed one k-regular network with k = 20 (see figure \ref{fig:network_properties_1_20regular} --and the corresponding entry on table \ref{tab:network_properties_merge_2_10regular}). The clustering coefficient of a 20-regular network is approximately the same as the one of a 10-regular network (higher than the combination of 10-regular networks). The average path length also decreases as there are more connections between the nodes. As we will show, these properties may vary. Since the nodes are subjected to random permutations, we can generate different overlaps that lead to distinct structural values.

Since we model the agent social space as a multitude of networks, we need to investigate the properties resulting from the merging of these networks. We do this empirically by taking the network instances used in the simulation and analysing the distribution of clustering coefficient and average path length for the different configurations.

\subsubsection{Properties of Overlapping K-Regular Networks}
\label{sec:overlapping_kreg}
\noindent We will now investigate what kind of properties we can get from merging k-regular networks. First, we analysed the distribution for the average path length and the clustering coefficient values over 100 network instances with 100 nodes each. From a box plot preliminary analysis (see figures \ref{append_fig:network_properties_cc_kreg} and \ref{append_fig:network_properties_apl_kreg}), we can see that the average path length does not vary much for the 100 instances, as such, we use the average as a descriptor for different configurations of k values and number of overlapping networks.

Figure~\ref{fig:network_properties_apl_line_kreg} shows the average value for the \textit{average path length} of 100 k-regular network instances, each with 100 nodes. We can see that the \textit{average path length} changes more drastically when we go from 1 to 2 networks. This is precisely the effect we can see in figure \ref{fig:network_properties_merge_2_10regular}. What we are seeing is that merging these networks at random effectively creates multiple shortcuts between points that were not connected in the original k-regular topology. Beyond this point, adding more networks does not modify this structural property in a significant way. Note that, since each network has 100 nodes, with $k=50$ the network is fully connected (hence the average path length being 1).

\begin{figure}[H]
	\centering
	\begin{subfigure}{.5\linewidth}
		\centering
		\includegraphics[width=1\linewidth]{"../analysis/pdf/network_properties_apl_line_kreg_12345"}
		\caption{}
		\label{fig:network_properties_apl_line_kreg_12345}
	\end{subfigure}%
	\begin{subfigure}{.5\linewidth}
		\centering
		\includegraphics[width=1\linewidth]{"../analysis/pdf/network_properties_apl_line_kreg_1020304050"}
		\caption{}
		\label{fig:network_properties_apl_line_kreg_1020304050}
	\end{subfigure}
	\begin{minipage}{0.9\textwidth}
		\vspace{0.2cm}
		\caption{Average value for the \textit{average path length} of 100 instances of overlapping k-regular networks (containing 100 nodes each) with k=\{1,2,3,4,5\}~(\ref{fig:network_properties_apl_line_kreg_12345}), and k= ~\{10,20,30,40,50\} (\ref{fig:network_properties_apl_line_kreg_1020304050}).}
		\label{fig:network_properties_apl_line_kreg}
	\end{minipage}
\end{figure}

\noindent Figure~\ref{fig:network_properties_cc_line_kreg} shows the results for the average \textit{clustering coefficient} of 100 instances of overlapping k-regular networks. We can see that adding multiple networks changes the clustering coefficient of the resulting structure. For $k \le 20$, the clustering coefficient drops when we add a second network. This happens because when we merge these networks, their connectivity is not high enough to generate shortcuts with tightly clustered neighbourhoods. The clustering coefficient is computed by taking the average local clustering coefficient of each node in the network. An initial highly clustered k-regular network (with all the nodes having the same structure) is affected when you modify the node neighbourhoods in such a way that some are more clustered than others.

@@

The exception is when you merge networks with $k=1$. These have a ring-like structure, so there are no nodes forming a clique\footnote{A clique is a group of nodes such that every two nodes are connected by an edge.} with their neighbours.

\subsubsection{Properties of Overlapping Scale-free Networks}
\noindent In a preliminary analysis we looked at the distribution of property values that results from merging scale-free networks (see figure \ref{append_fig:network_properties_sf}). Like in the previous results, the properties don't vary much between the 100 different instances of each overlapping network configuration. We can then say that specific configurations display a specific \textit{average path length} and \textit{clustering coefficient} both for \textit{k-regular} networks and \textit{scale-free} networks. Since the average of these measures is still a good descriptor, we plotted the average of the \textit{average path length} (figure \ref{fig:network_properties_apl_line_sf_12345}) and \textit{clustering coefficient} (figure \ref{append_fig:network_properties_cc_sf_12345}) for each overlapping network configuration.

@@

The parameter $d$ in figure \ref{fig:network_properties_line_sf} dictates how many edges are added by preferential attachment (see section\ref{sec:exp-setup_network_models}) each time a new node is added to a network (during its construction). This means that for $d=1$, the resulting network is a \textit{forest}: a network composed of disjoint tree graphs. Figure~\ref{fig:network_properties_apl_line_sf_12345} shows us that the \textit{average path length} still decreases consistently when we add more networks. One of the characteristics of \textit{scale-free} networks is its short average  path length, thus, for $d \ge 2$ we can see that adding more edges does not make so much difference. Scale-free networks also have a reduced \textit{clustering coefficient} which is evident in figure \ref{fig:network_properties_cc_line_sf_12345}. This is increased when we merge more networks.

\subsubsection{Merging K-Regular with Scale-Free Networks}
\noindent Finally, we look at what happens when we merge k-regular networks with \textit{scale-free} networks. We didn't make an exhaustive analysis for this configuration, but we show what is the resulting structure of such merging. You can see in figure \ref{fig:network_properties_merge_regular_scale-free} that this results in a network with a lower \textit{average path length} due to the shortcuts created by the \textit{scale-free} network. It is also less dense than the structure with two 10-regular networks (figure \ref{fig:network_properties_merge_2_10regular}). In this analysis we consider only the merging of two networks.

@@

Figure \ref{fig:network_properties_line_apl_sf_kreg} shows the \textit{average path length} of 100 independently generated instances of combinations of one \textit{k-regular} network and one \textit{scale-free} network with different values $k$ and $d$, respectively. We can see that for configurations where the \textit{k-regular} networks have low values of $k$ (figure \ref{fig:network_properties_apl_line_sf_kreg_12345}), adding a \textit{scale-free} network always decreases the \textit{average path length}. Moreover, the higher the $d$, the lower the path length, as more connections are formed between different nodes. With higher values of $k$ (figure \ref{fig:network_properties_apl_line_sf_kreg_1020304050}) however, increasing the value of $d$ does not decrease the \textit{average path length}. This is not surprising given that with that level of connectivity, having a more connected \textit{scale-free} network does not make such an impact. The first network is already highly dense and connected, so adding a few extra connections doesn't matter --what does matter in this case is that there is a scale-free network creating the extra shortcuts between nodes that were not connected before.

@@

\noindent Figure \ref{fig:network_properties_line_cc_sf_kreg} shows the results for the \textit{clustering coefficient}. These are similar to the results we get for \textit{k-regular} networks (see figure \ref{fig:network_properties_cc_line_kreg}). The difference is that we are not adding more networks but rather increasing the parameter $d$ of the scale-free. Curiously, in this case, the effects are qualitatively similar with the ones of merged \textit{k-regular} network with low value of $k$. With high values of $k$ (figure \ref{fig:network_properties_cc_line_sf_kreg_1020304050}) we can see that higher values of $d$ are not enough to make the network more clustered. This happens because scale-free networks are more sparse and have a very low number of triangles between nodes.

%***************************************************
%	Context Permeability 
%***************************************************
\subsection{Context Permeability}
\label{sec:results:context_permeability}

>>>>

        Since we construct scenarios where we overlap highly clustered k-regular networks and
        scale-free networks with node hubs, we are interested in the kind of properties that emerge
        from merging these structures -- and if/how they are comparable to the existing models
        --. The networks were produced by the \textit{b-have network library}
        \cite{Nunes:Software:11069}, exported to files, and analysed using the \textit{igraph R
          package}~\cite{igraph2006}.

@@

        % Overview
        \noindent In this section, we analyse and discuss the results for different experiments with
        both the context permeability and context switching models. We start by analysing the
        structural properties that arise from combining different networks. We then observe the
        impact of different model configurations in the convergence to consensus. We correlate the
        outcomes of our context permeability model simulations with the structural properties of the
        multi-network structures. We also explore how the new dynamics introduced by the context
        switching mechanism affects the consensus building process.

        \subsection{Overlapping Network Properties}
        \label{sec:network_properties}
        \noindent In this first analysis, we investigate the properties of different network
        topologies used in our models. One of the parameters in some experiments is the number of
        networks in which the agents interact. Adding more networks implies the addition of more
        connections. We show that what is important is \textit{not just} the the number of
        connections, but the properties of the resulting structure (when we merge the networks). We
        analyse the networks as follows.

        Each network is generated in such a way that the node indexes are randomised. This means
        that we can have multiple networks with the same topology and each node can have different
        neighbourhoods in different networks. Also, neighbourhoods do not necessarily overlap due to
        the node shuffling. Adding more networks to the social space is not the same as creating a
        single network with twice the number of connections.

        \noindent Consider figure \ref{fig:network_properties_merge_2_10regular}. We created two
        random k-regular networks with k=10 and merged them ignoring edges from common neighbours.
        (In the simulation models this edge agglutination is not performed, we use it here for the
        purposes of network analysis.) Common edges are highlighted with a different colour.

@@

        The networks are drawn using the \textit{Kamada-Kawai} Layout \cite{Kamada19897} which
        treats the ``geometric'' (Euclidean) distance between two vertices in the drawing as the
        ``graph theoretic'' distance between them in the corresponding graph. We can see by the
        results of the layout in figure \ref{fig:network_properties_merge_2_10regular}, that the
        distance between nodes in the network has decreased, this is also confirmed by the reduced
        average path length of the combined networks (see table
        \ref{tab:network_properties_merge_2_10regular}). To illustrate our point, we analysed one
        k-regular network with k = 20 (see figure \ref{fig:network_properties_1_20regular} and the
        corresponding entry on table \ref{tab:network_properties_merge_2_10regular}). The clustering
        coefficient of a 20-regular network is approximately the same as the one of a 10-regular
        network (higher than the combination of two 10-regular networks). The average path length
        also decreases as there are more connections between previously distant nodes. As we will
        show, these properties may vary due to the node indexes being subjected to random
        permutations.

        Since we model the agent social space as a multitude of networks, we need to investigate the
        properties resulting from the merging of these networks. We do this empirically by taking
        the network instances used in the simulations, and analysing the distribution of clustering
        coefficient and average path length for the different configurations.

        \subsubsection{Properties of Overlapping K-Regular Networks}
        \label{sec:overlapping_kreg}
        \noindent We will now investigate what kind of properties we can get from merging k-regular
        networks. First, we analysed the distribution for the average path length and the clustering
        coefficient values over 100 network instances (each with 100 nodes). From a box plot
        preliminary analysis (see figures \ref{append_fig:network_properties_cc_kreg} and
        \ref{append_fig:network_properties_apl_kreg}), we can see that the average path length does
        not vary much for the 100 instances, as such, the average makes a good descriptor for these
        properties.

        Figure~\ref{fig:network_properties_apl_line_kreg} shows the average value for the
        \textit{average path length} of 100 k-regular network instances (each with 100 nodes). We
        can see that the \textit{average path length} changes more drastically when we go from 1 to
        2 networks. This is precisely the effect we can see in figure
        \ref{fig:network_properties_merge_2_10regular}. Merging these networks at random effectively
        creates multiple shortcuts between points that were not connected in the original k-regular
        topology. Beyond this point, adding more networks does not modify this structural property
        in a significant way. Note that, since each network has 100 nodes, with $k=50$ the network
        is fully connected (hence the average path length being 1).

        \begin{figure}[H] \centering
          \begin{subfigure}{.5\linewidth} \centering
            \includegraphics[width=1\linewidth]{"../analysis/pdf/network_properties_apl_line_kreg_12345"}
            \caption{}
            \label{fig:network_properties_apl_line_kreg_12345}
          \end{subfigure}%
          \begin{subfigure}{.5\linewidth} \centering
            \includegraphics[width=1\linewidth]{"../analysis/pdf/network_properties_apl_line_kreg_1020304050"}
            \caption{}
            \label{fig:network_properties_apl_line_kreg_1020304050}
          \end{subfigure}

     \begin{minipage}{0.9\textwidth} \vspace{0.2cm}
            \caption{Average value for the \textit{average path length} of 100 instances of
              overlapping k-regular networks (containing 100 nodes each) with
              k=\{1,2,3,4,5\}~(\ref{fig:network_properties_apl_line_kreg_12345}), and k=
              ~\{10,20,30,40,50\} (\ref{fig:network_properties_apl_line_kreg_1020304050}).}
		\label{fig:network_properties_apl_line_kreg}
              \end{minipage}

        \end{figure}

        \noindent Figure~\ref{fig:network_properties_cc_line_kreg} shows the results for the average
        \textit{clustering coefficient} of 100 instances of overlapping k-regular networks. Adding
        multiple networks changes the clustering coefficient of the resulting structure. For
        $k \le 20$, the clustering coefficient drops when we add a second network. This happens
        because when we merge these networks, their connectivity is not high enough to generate
        shortcuts with tightly clustered neighbourhoods. The clustering coefficient is computed by
        taking the average local clustering coefficient of each node in the network. An initial
        highly clustered k-regular network (with all the nodes having the same structure) is
        affected when you modify the node neighbourhoods in such a way that some are more clustered
        than others. Networks with $k=1$ have a ring-like structure, so there are no nodes forming a
        clique\footnote{A clique is a group of nodes such that every two nodes are connected by an
          edge.} with their neighbours.

@@

        \subsubsection{Properties of Overlapping Scale-free Networks}
        \noindent We also looked at the properties that result from merging scale-free networks (see
        figure \ref{append_fig:network_properties_sf}). Like the previous results, the properties
        don't vary much between the 100 different instances. We can then say that specific
        configurations display a specific \textit{average path length} and \textit{clustering
          coefficient} both for \textit{k-regular} networks and \textit{scale-free} networks. We
        plotted the average of the \textit{average path length} (figure
        \ref{fig:network_properties_apl_line_sf_12345}) and \textit{clustering coefficient} (figure
        \ref{append_fig:network_properties_cc_sf_12345}) for each overlapping network configuration.

@@

        The parameter $d$ (figure \ref{fig:network_properties_line_sf}) dictates how many edges are
        added by preferential attachment (see section\ref{sec:exp-setup_network_models}). For $d=1$,
        the resulting network is a \textit{forest}: a network composed of disjoint tree
        graphs. Figure~\ref{fig:network_properties_apl_line_sf_12345} shows us that the
        \textit{average path length} still decreases consistently when we add more networks. One of
        the characteristics of \textit{scale-free} networks is its short average path length, thus,
        for $d \ge 2$, adding more edges does not make a significant difference. Scale-free networks
        also have a reduced \textit{clustering coefficient} (see figure
        \ref{fig:network_properties_cc_line_sf_12345}). When we add more networks the
        \textit{clustering coefficient} increases.

        \subsubsection{Merging K-Regular with Scale-Free Networks}
        \noindent Finally, we look at what happens when we merge k-regular networks with
        \textit{scale-free} networks. We didn't make an exhaustive analysis for this configuration,
        but we show what is the resulting structure of such merging. You can see in figure
        \ref{fig:network_properties_merge_regular_scale-free} that this results in a network with a
        lower \textit{average path length} due to the shortcuts created by the \textit{scale-free}
        network. It is also less dense than the structure with two 10-regular networks (figure
        \ref{fig:network_properties_merge_2_10regular}). We consider only the merging of two
        networks.

@@

        Figure \ref{fig:network_properties_line_apl_sf_kreg} shows the \textit{average path length}
        of 100 independently generated instances of combinations of one \textit{k-regular} network
        and one \textit{scale-free} network with different values $k$ and $d$ respectively. For
        configurations where the \textit{k-regular} networks have low values of $k$ (figure
        \ref{fig:network_properties_apl_line_sf_kreg_12345}), adding a \textit{scale-free} network
        always decreases the \textit{average path length}. Moreover, the higher the $d$, the lower
        the path length (as more connections are formed between different nodes). With higher values
        of $k$ (figure \ref{fig:network_properties_apl_line_sf_kreg_1020304050}) however, increasing
        the value of $d$ does not decrease the \textit{average path length}. This is not surprising
        given the level of connectivity of this structure. The first network is already highly dense
        and connected, so adding a few extra connections doesn't matter; what does matter is that we
        have a scale-free network creating shortcuts between nodes that were not previously
        connected.

@@

        \noindent Figure \ref{fig:network_properties_line_cc_sf_kreg} shows the results for the
        \textit{clustering coefficient}. These are similar to the results we get for
        \textit{k-regular} networks (see figure \ref{fig:network_properties_cc_line_kreg}). The
        difference is that we are not adding more networks but rather increasing the parameter $d$
        of the scale-free (number of connections added by preferential attachment for each
        note). In this case, the effects are qualitatively similar with the ones of merged
        \textit{k-regular} network with low value of $k$. In contrast, for high values of $k$
        (figure \ref{fig:network_properties_cc_line_sf_kreg_1020304050}), a higher value of $d$ is
        not enough to make the network more clustered. This happens because scale-free networks are
        more sparse and have a very low number of triangles between nodes.

        % *************************************************** 
        %           Context Permeability
        % ***************************************************
        \subsection{Context Permeability}
\label{sec:results:context_permeability}
